---
title: "Chi_Sq_lifeform_vs_biovigilance"
format: html
---

### Setting up the Document:

Here, I will load the data I am performing a chi-square test of independence on the variables lifeform and biovigilance.

```{r}
rm(list=ls())
library(here)
library(ggplot2)
library(ggfortify)
weeds<- read.csv(here("Data", "weeds_clean.csv"))
weeds$biovigilance<- as.factor(weeds$biovigilance)
weeds$lifeform<- as.factor(weeds$lifeform)
#correcting varibles identities 
glimpse(weeds)
```

This dataset contains 1,294 rows and 5 columns, capturing information on lifeforms and their classifications, with some missing values (NA). The red_UK column represents conservation statuses, now a factor with three levels: NT, VU, and Wait. The lifeform column categorizes observations by biological form, now a factor with five levels: Therophyte, Hemicryptophyte, Geophyte, Phanerophyte, and Liana. The biogeo column indicates the biogeographical region, now a factor with four levels: Cosmopolitan, Eurasian, European, and Mediterranean. The module column, originally an integer, is now a factor with four levels: 1, 3, 5, and 99, corresponding to habitat categories. Finally, biovigilance is now a factor with two levels: 1 (present) and 0 (absent). This dataset combines categorical and numeric data, providing a foundation for exploring patterns in lifeform classifications, conservation statuses, geographic distributions, and monitoring efforts.

### Analysis

Null Hypothesis:
There is no significant association between lifeform and biovigilance. The observed distribution of lifeforms across biovigilance categories is due to chance and matches the expected distribution under independence.

Alternative Hypothesis:
There is a significant association between lifeform and biovigilance. The observed distribution of lifeforms across biovigilance categories differs significantly from the expected distribution under independence.

```{r}
totals<- weeds|> group_by(lifeform, biovigilance)|>
  filter(!is.na(biovigilance) & !is.na(red_UK)) |> # Remove rows with NA in 'lifeform' or 'red_UK'
  summarise(total.numbers = n(), .groups = 'drop' )
```

NA values in lifeform and biovigilance created additional, unintended groups in the analysis, as NA is treated as a separate category. These artificial groups could distort results and misrepresent the true relationship between the variables. Removing rows with NA ensures a cleaner, more accurate visualization that reflects only meaningful data.

```{r}
ggplot(totals, aes(x = lifeform, y = total.numbers, fill = biovigilance)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "Distribution of Lifeforms by Red_UK",
    x = "Lifeform",
    y = "Total Numbers",
    fill = "Red_UK Status"
  ) +
  theme_minimal()+
  coord_flip()
```

The chart shows that therophytes and hemicryptophytes are the most common lifeforms, with high numbers across both Red_UK statuses, though therophytes show slightly more balance between statuses 0 and 1. Geophytes and chamaephytes have fewer observations, with chamaephytes almost entirely associated with Red_UK status 0. Phanaerophytes and lianas are rare, with very few observations, while hydrophytes show a small but relatively balanced distribution between statuses 0 and 1. This highlights clear differences in the prevalence and monitoring of different lifeforms.


$Chi^2$ Test
```{r}
# Create a contingency table for lifeform and biovigilance
lady.mat<- xtabs(~lifeform + biovigilance, data=weeds)
```

```{r}
lady.chi<-chisq.test(lady.mat)
lady.chi
```

A Pearson’s Chi-Squared test of independence was performed to assess the relationship between lifeform and biovigilance. The results showed a significant association between the two variables, χ²(7, N = total count) = 19.008, p = 0.008. This indicates that the distribution of lifeforms is not independent of their biovigilance status, as the observed counts differ significantly from the expected counts under the null hypothesis.

```{r}
ggplot(totals, aes(x = lifeform, y = total.numbers, fill = biovigilance)) +
  geom_bar(stat = "identity", position = "dodge", color = "black", linewidth = 0.2) +
  scale_fill_brewer(palette = "Set2", name = "Biovigilance Status") +
  labs(
    title = "Distribution of Lifeforms by Biovigilance",
    x = "Lifeform",
    y = "Total Numbers",
    fill = "Biovigilance Status"
  ) +
  theme_minimal() +
  coord_flip()+
  theme(plot.title = element_text(face = "bold"),
        legend.position = "top")

```


Rejecting the Null Hypothesis:
Based on the chi-square test results, we reject the null hypothesis, as the p-value (0.008) is below the standard significance level (e.g., 0.05). This indicates a significant association between lifeform and biovigilance status, meaning the observed differences in the distribution of biovigilance across lifeforms are unlikely to be due to chance and reflect a meaningful relationship.
