---
title: "Chi_Sq_redUK_vs_lifeform"
format: html
---

### Setting up the Document:

Here, I will load the data I am performing a chi-square test of independence on the variables lifeform and red_UK.

```{r}
rm(list=ls())
library(here)
library(ggplot2)
library(ggfortify)
weeds<- read.csv(here("Data", "weeds_clean.csv"))
weeds$red_UK<- as.factor(weeds$red_UK)
#chi square works when both are characters or only when red_UK is fct it won't work if lifeform is a fct the chi sq value comes out as an NA
glimpse(weeds)
```

This dataset contains 1,294 rows and 5 columns, capturing information on lifeforms and their classifications, with some missing values (NA). The red_UK column represents conservation statuses, now a factor with three levels: NT, VU, and Wait. The lifeform column categorizes observations by biological form, now a factor with five levels: Therophyte, Hemicryptophyte, Geophyte, Phanerophyte, and Liana. The biogeo column indicates the biogeographical region, now a factor with four levels: Cosmopolitan, Eurasian, European, and Mediterranean. The module column, originally an integer, is now a factor with four levels: 1, 3, 5, and 99, corresponding to habitat categories. Finally, biovigilance is now a factor with two levels: 1 (present) and 0 (absent). This dataset combines categorical and numeric data, providing a foundation for exploring patterns in lifeform classifications, conservation statuses, geographic distributions, and monitoring efforts.

### Analysis

Null Hypothesis:
There is no significant association between lifeform and Red_UK status. The observed distribution of lifeform across Red_UK categories is due to chance and matches the expected distribution under independence.

Alternative Hypothesis:
There is a significant association between lifeform and Red_UK status. The observed distribution of lifeform across Red_UK categories differs significantly from the expected distribution under independence.

```{r}
totals<- weeds|> group_by(lifeform, red_UK)|>
  filter(!is.na(lifeform) & !is.na(red_UK)) |> # Remove rows with NA in 'lifeform' or 'red_UK'
  summarise(total.numbers = n(), .groups = 'drop' )
```

NA values in lifeform and red_UK created additional, unintended groups in the analysis, as NA is treated as a separate category. These artificial groups could distort results and misrepresent the true relationship between the variables. Removing rows with NA ensures a cleaner, more accurate visualization that reflects only meaningful data.

```{r}
ggplot(totals, aes(x = lifeform, y = total.numbers, fill = red_UK)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "Distribution of Lifeforms by Red_UK",
    x = "Lifeform",
    y = "Total Numbers",
    fill = "Red_UK Status"
  ) +
  theme_minimal()+
  coord_flip()
```

The distribution of lifeforms across Red_UK statuses shows a notable imbalance, with certain lifeforms dominating the dataset. For example, therophytes and hemicryptophytes exhibit significantly higher counts compared to other lifeforms, indicating their prevalence in the dataset. In contrast, categories like liana and hydrophyte have minimal representation, suggesting their relative rarity. Within each lifeform, the Red_UK statuses are also unevenly distributed, with certain statuses like LC (Least Concern) dominating most groups, reflecting potential patterns in conservation status classification

$Chi^2$ Test
```{r}
# Create a contingency table for lifeform and red_UK
lady.mat<- xtabs(~lifeform + red_UK, data=weeds)
```

```{r}
lady.chi<-chisq.test(lady.mat)
lady.chi
```

A Pearson’s Chi-Squared test of independence was performed to assess the relationship between lifeform and Red_UK status. The results showed no significant association between the two variables, χ²(42, N = total count) = 49.42, p = 0.201. This indicates that the distribution of lifeforms is independent of their Red_UK status, as the observed counts align closely with the expected counts under the null hypothesis.

```{r}
ggplot(totals, aes(x = lifeform, y = total.numbers, fill = red_UK)) +
 geom_bar(stat = "identity", position = "dodge", color = "black", linewidth = 0.2) +
  labs(
    title = "Distribution of Lifeforms by Red_UK",
    x = "Lifeform",
    y = "Total Numbers",
    fill = "Red_UK Status"
  ) +
  scale_fill_brewer(palette = "Set3", name = "red_UK") +
  theme_minimal()+
  coord_flip()+
  theme(plot.title = element_text(face = "bold"))
```

Failing to Reject the Null Hypothesis:
Based on the chi-square test results, we fail to reject the null hypothesis, as the p-value (0.201) exceeds the standard significance level (e.g., 0.05). This suggests there is no significant association between lifeform and Red_UK status, meaning the observed differences in the distribution of statuses across lifeforms could be attributed to chance rather than a meaningful relationship.

