---
title: "Chi_Sq_module_vs_biogeo"
format: html
---


### Setting up the Document:

Here, I will load the data I am performing a chi-square test of independence on the variables biogeo and module

```{r}
rm(list=ls())
library(here)
library(ggplot2)
library(ggfortify)
weeds<- read.csv(here("Data", "weeds_clean.csv"))
weeds$biogeo<- as.factor(weeds$biogeo)
weeds$module<- as.factor(weeds$module)
#correcting varibles identities 
glimpse(weeds)
```

This dataset contains 1,294 rows and 5 columns, capturing information on lifeforms and their classifications, with some missing values (NA). The red_UK column represents conservation statuses, now a factor with three levels: NT, VU, and Wait. The lifeform column categorizes observations by biological form, now a factor with five levels: Therophyte, Hemicryptophyte, Geophyte, Phanerophyte, and Liana. The biogeo column indicates the biogeographical region, now a factor with four levels: Cosmopolitan, Eurasian, European, and Mediterranean. The module column, originally an integer, is now a factor with four levels: 1, 3, 5, and 99, corresponding to habitat categories. Finally, biovigilance is now a factor with two levels: 1 (present) and 0 (absent). This dataset combines categorical and numeric data, providing a foundation for exploring patterns in lifeform classifications, conservation statuses, geographic distributions, and monitoring efforts.

### Analysis

Null Hypothesis:
There is no significant association between biogeo and module. The observed distribution of biogeo categories across module levels is due to chance and matches the expected distribution under independence.

Alternative Hypothesis:
There is a significant association between biogeo and module. The observed distribution of biogeo categories across module levels differs significantly from the expected distribution under independence.

```{r}
totals<- weeds|> group_by(module, biogeo)|>
  filter(!is.na(biogeo) & !is.na(module)) |> # Remove rows with NA in 'lifeform' or 'red_UK'
  summarise(total.numbers = n(), .groups = 'drop' )
```

NA values in module and biogeo created additional, unintended groups in the analysis, as NA is treated as a separate category. These artificial groups could distort results and misrepresent the true relationship between the variables. Removing rows with NA ensures a cleaner, more accurate visualization that reflects only meaningful data.

```{r}
ggplot(totals, aes(x = module, y = total.numbers, fill = biogeo)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "Distribution of Module by Biogeo",
    x = "Module",
    y = "Total Numbers",
    fill = "Biogeo"
  ) +
  theme_minimal()+
  coord_flip()
```
The distribution of module by biogeo reveals distinct patterns. Module 5 is heavily dominated by the Mediterranean biogeographical region, contributing significantly more observations compared to other regions. Other modules, such as 1, 3, 9 and 99, show more balanced distributions across biogeographical categories like Cosmopolitan, Eurasian, and European, but with fewer total observations. Module 9 has relatively few entries, with a noticeable presence of European and Eurasian regions. Overall, the Mediterranean region appears to have a disproportionate influence on Module 5, while other regions are more evenly distributed across the remaining modules.


$Chi^2$ Test
```{r}
# Create a contingency table for biogeo and module
lady.mat<- xtabs(~module + biogeo, data=weeds)
```

```{r}
lady.chi<-chisq.test(lady.mat)
lady.chi
```
A Pearson’s Chi-Squared test of independence was conducted to assess the relationship between biogeo and module. The results showed a highly significant association between the two variables, χ²(12, N = total count) = 285.08, p < 2.2e-16. This indicates that the distribution of biogeo categories is not independent of module, suggesting a meaningful relationship between biogeographical regions and module classifications.


```{r}
library(ggplot2)

ggplot(totals, aes(x = module, y = total.numbers, fill = biogeo)) +
  geom_bar(stat = "identity", position = "dodge", color = "black", linewidth = 0.2) + 
  scale_fill_brewer(palette = "Set3", name = "Biogeo") + 
  labs(
    title = "Distribution of Module by Biogeo",
    x = "Module",
    y = "Total Numbers",
    fill = "Biogeographical Region"
  ) +
  theme_minimal() +
  coord_flip()+
  theme(legend.position = "top",
        plot.title = element_text(face = "bold"))

```


Rejecting the Null Hypothesis:
Based on the chi-square test results, we reject the null hypothesis, as the p-value (< 2.2e-16) is well below the standard significance level (e.g., 0.05). This indicates a significant association between biogeo and module, meaning the observed differences in the distribution of biogeo categories across module levels are unlikely to be due to chance and reflect a meaningful relationship.
