---
title: "Chi_Sq_lifeform_vs_biovigilance"
format: html
---

### Setting up the Document:

Here, I will load the data I am performing a chi-square test of independence on the variables lifeform and biovigilance.

```{r}
rm(list=ls())
library(here)
library(tidyverse)
library(ggplot2)
library(ggfortify)
weeds<- read.csv(here("Data", "weeds_clean.csv"))
weeds$biovigilance<- as.factor(weeds$biovigilance)
weeds$lifeform<- as.factor(weeds$lifeform)
#correcting varibles identities 
glimpse(weeds)
```

The dataset contains 1,294 rows and 5 columns, with all variables now represented as categorical factors. The red_UK column, indicating conservation statuses, has three levels: NT (Near Threatened), VU (Vulnerable), and Wait. The lifeform column categorizes observations into five biological forms: Therophyte, Hemicryptophyte, Geophyte, Phanerophyte, and Liana. The biogeo column reflects the biogeographical region, with four levels: Cosmopolitan, Eurasian, European, and Mediterranean. The module column, initially an integer, now represents habitat categories as a factor with four levels: 1, 3, 5, and 99. Finally, the biovigilance column is a binary factor with two levels: 1 (present) and 0 (absent). Contrary to the initial interpretation, this dataset contains only categorical data, not a mix of categorical and numeric variables. This provides a robust framework for analyzing patterns in lifeform classifications, conservation statuses, geographic distributions, and monitoring efforts using appropriate categorical data techniques.

### Analysis

Null Hypothesis:
There is no significant association between lifeform and biovigilance. The observed distribution of lifeforms across biovigilance categories is consistent with the expected distribution under the assumption of independence.

Alternative Hypothesis:
There is a significant association between lifeform and biovigilance. Specifically, certain lifeforms, such as Therophytes (short-lived plants that are highly responsive to environmental changes), may be overrepresented in biovigilance categories due to their detectability and ecological traits. Conversely, lifeforms like Phanerophytes (trees and shrubs with longer lifespans) may be underrepresented, potentially due to lower detection rates or monitoring biases.

```{r}
totals <- weeds |> group_by(lifeform, biovigilance) |>
  filter(
    !is.na(biovigilance) & (lifeform != "")
) |>
  summarise(total.numbers = n())
```

NA values in lifeform and biovigilance created additional, unintended groups in the analysis, as NA is treated as a separate category. These artificial groups could distort results and misrepresent the true relationship between the variables. Removing rows with NA ensures a cleaner, more accurate visualization that reflects only meaningful data.

```{r}
ggplot(totals, aes(x = lifeform, y = total.numbers, fill = biovigilance)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "Distribution of Lifeforms by Red_UK",
    x = "Lifeform",
    y = "Total Numbers",
    fill = "Red_UK Status"
  ) +
  theme_minimal()+
  coord_flip()
```

The chart shows that therophytes are the most common lifeform, with a high number of observations across both Red_UK statuses, though status 0 is significantly more prevalent than status 1. Hemicryptophytes are the second most common, primarily associated with status 0, with smaller proportions for status 1.

Geophytes and chamaephytes have fewer observations, with chamaephytes almost entirely associated with status 0. Phanaerophytes and lianas are rare, with very few observations, while hydrophytes exhibit a small but relatively balanced distribution between statuses 0 and 1.

These patterns reveal clear differences in the prevalence and monitoring of different lifeforms, highlighting potential ecological or observational factors influencing their association with the Red_UK statuses.

$Chi^2$ Test
```{r}
# Create a contingency table for lifeform and biovigilance
lady.mat<- xtabs(~lifeform + biovigilance, data=weeds)
```

```{r}
lady.chi<-chisq.test(lady.mat)
lady.chi
```

A Pearson’s Chi-Squared test of independence was performed to assess the relationship between lifeform and biovigilance. The results showed a significant association between the two variables, χ²(7, N = total count) = 19.008, p = 0.008. This indicates that the distribution of lifeforms is not independent of their biovigilance status, as the observed counts differ significantly from the expected counts under the null hypothesis.

```{r}
ggplot(totals, aes(x = lifeform, y = total.numbers, fill = biovigilance)) +
  geom_bar(stat = "identity", position = "dodge", color = "black", linewidth = 0.2) +
  scale_fill_brewer(palette = "Set2", name = "Biovigilance Status") +
  labs(
    title = "Distribution of Lifeforms by Biovigilance",
    x = "Lifeform",
    y = "Total Numbers",
    fill = "Biovigilance Status"
  ) +
  theme_minimal() +
  coord_flip()+
  theme(plot.title = element_text(face = "bold"),
        legend.position = "top")

```


Rejecting the Null Hypothesis:
Based on the chi-square test results, we reject the null hypothesis, as the p-value (0.008) is below the standard significance level of 0.05. This indicates a significant association between lifeform and Red_UK status, meaning the observed differences in the distribution of Red_UK status across lifeforms are unlikely to be due to chance and reflect a meaningful relationship.

## Post Hoc Analysis
```{r}
library(chisq.posthoc.test)

posthoc<-chisq.posthoc.test(lady.mat, method = "bonferroni")
posthoc
```

The post hoc analysis of lifeform versus biovigilance reveals no significant associations between lifeform and biovigilance status, as indicated by the p-values, which are all greater than the standard significance threshold of 0.05. Biovigilance is coded as 0 and 1, with 0 representing species that are not under active biovigilance or monitoring for conservation purposes, and 1 representing species that are being actively monitored due to their conservation concern. The residuals indicate the differences between the observed and expected values for each lifeform under the two biovigilance categories.

For chamaephyte, geophyte, hydrophyte, liana, phanaerophyte, and therophyte, the residuals are either positive or negative, suggesting some variation in their presence across the biovigilance categories. However, the p-values for all these comparisons are above 0.05, suggesting that these differences are not statistically significant. This means that the distribution of these lifeforms between the 0 and 1 categories of biovigilance is likely due to random variation rather than a meaningful ecological pattern.

The hemicryptophyte lifeform, with residuals close to zero for both categories, also shows p-values of 1, reinforcing that there is no significant relationship between this lifeform and its inclusion in biovigilance monitoring.

In conclusion, the lack of significant results from the post hoc test implies that there is no clear distinction between species lifeforms in terms of their biovigilance status (0 = not monitored, 1 = monitored). It suggests that factors other than lifeform—such as population size, habitat type, or ecological traits—might better explain the variation in biovigilance categorization rather than lifeform alone.

The chi-square test may have returned a significant result despite no meaningful differences due to factors like small sample sizes or low expected frequencies in certain categories, which can inflate the test statistic. Additionally, the chi-square test does not account for potential confounding variables, which could result in spurious significance. While the chi-square test indicated significance, the post hoc pairwise tests with Bonferroni correction revealed no significant differences, suggesting that the chi-square result might be influenced by random variation, multiple comparisons, or data structure, rather than reflecting a true biological relationship.






