---
title: "Chi_Sq_redUK_vs_lifeform"
format: html
---

### Setting up the Document:

Here, I will load the data I am performing a chi-square test of independence on the variables lifeform and red_UK.

```{r}
rm(list=ls())
library(tidyverse)
library(here)
library(ggplot2)
library(ggfortify)
weeds<- read.csv(here("Data", "weeds_clean.csv"))
weeds$red_UK<- as.factor(weeds$red_UK)

#chi square works when both are characters or only when red_UK is fct it won't work if lifeform is a fct the chi sq value comes out as an NA
glimpse(weeds)
```

This dataset contains 1,294 rows and 5 columns, capturing information on lifeforms and their classifications, with some missing values (NA). The red_UK column represents conservation statuses, now a factor with three levels: NT, VU, and Wait. The lifeform column categorizes observations by biological form, now a factor with five levels: Therophyte, Hemicryptophyte, Geophyte, Phanerophyte, and Liana. The biogeo column indicates the biogeographical region, now a factor with four levels: Cosmopolitan, Eurasian, European, and Mediterranean. The module column, originally an integer, is now a factor with four levels: 1, 3, 5, and 99, corresponding to habitat categories. Finally, biovigilance is now a factor with two levels: 1 (present) and 0 (absent). This dataset combines categorical and numeric data, providing a foundation for exploring patterns in lifeform classifications, conservation statuses, geographic distributions, and monitoring efforts.

### Analysis

Null Hypothesis:
 There is no significant association between lifeform and Red_UK status. The observed distribution of lifeforms across Red_UK categories is consistent with what would be expected under conditions of ecological independence, implying that lifeform types do not influence or are not influenced by conservation status. This would suggest that different lifeforms, regardless of their biological characteristics, are equally likely to be classified into various Red_UK conservation statuses.
 
Alternative Hypothesis:
The distribution of lifeforms across Red_UK statuses differs significantly, implying that lifeforms with particular life history traits (such as slower reproduction rates, longer lifespans, or specialized habitat requirements) are more prone to being categorized under higher conservation concern statuses. For example, phanerophytes (woody plants) might be more likely to be categorized as VU due to their slow growth and limited dispersal ability, making them more susceptible to environmental pressures.


```{r}
totals<- weeds|> group_by(lifeform, red_UK)|>
  filter(!is.na(lifeform) & !is.na(red_UK)) |> # Remove rows with NA in 'lifeform' or 'red_UK'
  summarise(total.numbers = n(), .groups = 'drop' )

```

NA values in lifeform and red_UK created additional, unintended groups in the analysis, as NA is treated as a separate category. These artificial groups could distort results and misrepresent the true relationship between the variables. Removing rows with NA ensures a cleaner, more accurate visualization that reflects only meaningful data.

```{r}

ggplot(totals, aes(x = lifeform, y = total.numbers, fill = red_UK)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "Distribution of Lifeforms by Red_UK",
    x = "Lifeform",
    y = "Total Numbers",
    fill = "Red_UK Status"
  ) +
  theme_minimal()+
  coord_flip()
```
The distribution of lifeforms across Red_UK statuses reveals a notable imbalance, with certain lifeforms dominating the dataset. For instance, therophytes (annual plants) and hemicryptophytes (perennial herbs) exhibit significantly higher counts compared to other lifeforms, suggesting their relative abundance in the data. In contrast, categories like lianas (climbing plants) show minimal representation, indicating their rarity in this dataset. Notably, the geophytes (underground storage plants) and phanerophytes (woody plants) also show intermediate frequencies.

Therophytes, characterized by their rapid life cycles and dependence on specific, often disturbed habitats, may show a higher vulnerability to environmental stressors. These plants typically thrive in environments with seasonal or periodic disturbances, such as deserts or wetland ecosystems. However, their short lifespan and dependence on favorable conditions for germination and growth make them sensitive to climate changes, habitat destruction, and other ecological threats. This vulnerability may explain the observed spike in "VU" (Vulnerable) status among therophytes, as their specialized habitats and inability to adapt quickly to environmental changes may put them at greater risk compared to more resilient lifeforms.

Within each lifeform, the distribution of Red_UK statuses is uneven, with certain statuses like LC (Least Concern) predominantly represented in most groups. This imbalance hints at possible conservation patterns, with some lifeforms being more commonly classified as LC while others may be more frequently categorized under other conservation statuses such as VU (Vulnerable) or NT (Near Threatened). The higher frequency of therophytes in the "VU" category suggests that lifeforms with specific ecological traits, such as limited dispersal abilities or narrow habitat preferences, are more likely to be assigned higher conservation concern statuses.

Such trends likely reflect inherent conservation risks tied to the ecological characteristics and life history traits of different plant types. Lifeforms with more specialized needs or those that face significant environmental pressures may be more prone to classification under higher conservation concern statuses, highlighting the need for targeted conservation efforts for these vulnerable species.

$Chi^2$ Test

```{r}
# Create a contingency table for lifeform and red_UK
lady.mat<- xtabs(~lifeform + red_UK, data=weeds)
```

```{r}
lady.chi<-chisq.test(lady.mat)
lady.chi
```

A Pearson’s Chi-Squared test of independence was performed to assess the relationship between lifeform and Red_UK status. The results showed no significant association between the two variables, χ²(42, N = total count) = 49.42, p = 0.201. This indicates that the distribution of lifeforms is independent of their Red_UK status, as the observed counts align closely with the expected counts under the null hypothesis.

```{r}
# Grouping low count levels together for chi-square analysis
weeds$red_UK_combined <- weeds$red_UK  # Create a new variable for the combined Red_UK status

# Combine "Data Deficient" and "Least Concern" for geophytes
weeds$red_UK_combined <- ifelse(weeds$lifeform == "Geophyte" & weeds$red_UK %in% c("DD", "LC"),
                                "LC/DD", weeds$red_UK_combined)

# Update the contingency table to reflect the new grouping
lady.mat_combined <- xtabs(~lifeform + red_UK_combined, data = weeds)

# Perform chi-square test on the grouped data
lady.chi_combined <- chisq.test(lady.mat_combined)
lady.chi_combined
#give the same output did I do it wrong                  
```


```{r}
ggplot(totals, aes(x = lifeform, y = total.numbers, fill = red_UK)) +
 geom_bar(stat = "identity", position = "dodge", color = "black", linewidth = 0.2) +
  labs(
    title = "Distribution of Lifeforms by Red_UK",
    x = "Lifeform",
    y = "Total Numbers",
    fill = "Red_UK Status"
  ) +
  scale_fill_brewer(palette = "Set3", name = "red_UK") +
  theme_minimal()+
  coord_flip()+
  theme(plot.title = element_text(face = "bold"))
```

Failing to Reject the Null Hypothesis:
Based on the chi-square test results, we fail to reject the null hypothesis, as the p-value (0.201) exceeds the standard significance level of 0.05. This suggests that, after grouping lifeforms with low counts together, there is no significant association between lifeform and Red_UK status. The observed differences in the distribution of Red_UK statuses across lifeforms can likely be attributed to random variation, rather than reflecting a meaningful ecological relationship. Therefore, we conclude that the distribution of lifeforms across conservation statuses does not significantly differ from what would be expected under independence.

## Post Hoc
```{r}
library(chisq.posthoc.test)

posthoc<-chisq.posthoc.test(lady.mat, method = "bonferroni")
posthoc
```

p value 0.00034300 for LC and hemicryptophyte
0.0766090 pvalue for vu and hemicryptophyte
0.0520270 pvalue for en and therophyte
0.0075860 pvalue for  VU and therophyte

The post hoc pairwise proportions test results reveal significant and borderline associations between certain lifeforms and Red_UK conservation statuses, providing insights into their ecological characteristics and vulnerabilities. For instance, the strong association between hemicryptophytes (perennial herbs) and the LC (Least Concern) status (p = 0.000343) suggests that these plants, often found in a variety of habitats with relatively stable populations, are less susceptible to environmental threats, making them more likely to be classified as Least Concern. However, the borderline p-value (0.0766) for hemicryptophytes and the VU (Vulnerable) status suggests a weak trend that, while not statistically significant at the conventional threshold, may indicate that some hemicryptophytes are at risk in certain environments, potentially due to habitat fragmentation or other localized threats. Similarly, the borderline p-value (0.052) for therophytes (annual plants) and the EN (Endangered) status implies that these short-lived, fast-reproducing plants could be vulnerable to environmental disturbances such as climate change, which may lead to their increased risk of endangerment. The significant association between therophytes and VU (Vulnerable) (p = 0.007586) further supports the idea that their rapid reproduction, while advantageous under stable conditions, might not be enough to counteract the pressures of habitat loss and other ecological threats, making them more likely to be classified as vulnerable. These results underscore how life history traits—such as longevity, reproductive strategy, and habitat requirements—play a crucial role in determining a species' conservation status, with certain lifeforms being more susceptible to environmental pressures and thus requiring focused conservation efforts.

